---
title: "Data Dictionary"
subtitle: APRA Data Science Courses
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

## Introduction

This project is available on [GitHub](https://github.com/majerus/apra_data_science_courses).

fictious data, produced by generate R script

Challenges built in

## Biographical Data {.tabset}

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error=FALSE)

```

```{r}
# load tidyverse library 
library(tidyverse)
library(knitr)
library(plotly)
library(DT)

# read in biographical data from .csv file on github into a data frame in R called bio
# bio <- read_csv("https://raw.githubusercontent.com/majerus/apra_data_science_courses/master/bio_data_table.csv")
bio <- read_csv("bio_data_table.csv")
```

The biographical data has `r ncol(bio)` variables and `r nrow(bio)` observations. The data is stored at the donor level. Each row of the data represents a unique donor and biographical information about that donor.  

### Numeric Variables

```{r, echo=FALSE}
# select only numeric variables from bio
numeric_variables <- 
  bio %>% 
  select_if(is.numeric) 
```

There are `r ncol(numeric_variables)` numeric variables:

- **id**: A seven digit numeric id that is unique to each donor.  
- **household_id**: A seven digit numeric id that is unique to households. More than one donor may share a household_id. 
- **lat**: The latitude of the center point of each donor's zipcode. Missing for donor's residing outside the United States.  
- **lon**: The longitude of the center point of each donor's zipcode. Missing for donor's residing outside the United States.

```{r}
# select and glimpse only numeric variables
bio %>% 
  select_if(is.numeric) %>% 
  glimpse()

```

### Character Variables {.tabset .tabset-fade .tabset-pills}

```{r, echo=FALSE}
# select only numeric variables from bio
character_variables <- 
  bio %>% 
  select_if(is.character) 
```

When loaded by default there are `r ncol(character_variables)` character variables:

- name: Each donor's first and last name formatted as "last name, first name".
- country: Each donor's country of residence. 
- city: Each donor's city of residence. 
- deceased: A binary indicator that indicates if a donor is deceased ("Y"|"N")
- zip: The five digit zipcode of donor's whose country of residence is the United States. 
- state: The two-letter state abbreviation for each donor whose country of residence is the United States. 
- capacity: Each donor's capacity represented within an estimated range. 
- capacity_source: A categorical variable indicating how the capacity was determined ("institutional"|"screening").
- race: a categorical variable indicating the donor's race. 

```{r}
# select and glimpse only character variables
bio %>% 
  select_if(is.character) %>%  
  glimpse()

```

```{r, echo=FALSE,include = FALSE}
# You need this code to conduct the magic dependences attaching...
# I usually forget this!!! But it is needed for generating DT tables in loops, etc. 
datatable(NULL)
plot_ly(mtcars)
```

```{r results='asis'}
# select character variables to be included in tables and plots
chars <- 
  bio %>% 
  select_if(is.character) %>% 
  select(-name, -city, -zip) %>% 
  colnames() 

# loop to create table and plot of counts of each character variable
# could replace this with purrr
for (i in chars) {
  
  current <- character_variables[, i]
  
  cat(paste("\n\n####", i, "\n"))
  
  print(htmltools::tagList(
    current %>%
      count(.data[[i]]) %>%
      datatable(rownames = FALSE),
    
    ggplotly(
      ggplot(current, aes_string(x = i)) +
        geom_bar(fill = '#1f77b4') +
        theme_minimal())
    
  ))
  
}


```

### Date Variables {.tabset .tabset-fade .tabset-pills}

```{r, echo=FALSE}
# select only numeric variables from bio
date_variables <- 
  bio %>% 
  select_if(function(x) inherits(x, 'Date'))
```

There is `r ncol(date_variables)` date variable:

- birthday: The date of each donor's birth stored as a date variable.

```{r}
# select and glimpse only date variable(s)
bio %>% 
  select_if(function(x) inherits(x, 'Date')) %>% 
  glimpse()

```

```{r results='asis'}
# select character variables to be included in tables and plots
ggplotly(
  ggplot(date_variables, aes(x = birthday)) +
    geom_histogram(fill = '#1f77b4') +
    theme_minimal())


```
