---
title: "Example Exercises"
date: 6/9/2020
output:
 html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
# set rmarkdown options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# load R packages
library(tidyverse)
library(janitor)
library(lubridate)
library(plotly)
library(DT)
library(leaflet)
library(scales)
```

# Load Data {.tabset}

First, let's take a look at the data that is available to us. 

## Loading data from .csv files

**How do you read a .csv into R?**

```{r load_data}
# read in biographical data table
bio <- read_csv("https://raw.githubusercontent.com/majerus/apra_data_science_courses/master/bio_data_table.csv")

# read in giving data table
giving <- read_csv("https://raw.githubusercontent.com/majerus/apra_data_science_courses/master/giving_data_table.csv")

```

You can read multiple data files into the same R session. Each of these files contain fictional data created by the `generate_data.R` script. 

## Loading data from a database

**Have you ever connected R to a database?**

The following is an example of how to create a sample database in R and to load information from that database. More information on using databases from R can be found [here](https://db.rstudio.com/). If you use the tidyverse, you can use the same workflow with information from databases and .csv or Excel files. 

```{r}
# create database connection
con <- DBI::dbConnect(RSQLite::SQLite(), path = ":dbname:")

# put some data in our new database
copy_to(dest = con,
        df = bio,
        name = "bio_table",
        temporary = FALSE)

copy_to(dest = con,
        df = giving ,
        name = "giving_table",
        temporary = FALSE)

# print out our table names
db_list_tables(con)

# let's take a look at the bio table
tbl(con, "bio_table") 

# we can use dplyr syntax to query a database
# dplyr automatically converts our r code to sql
# alternatively you can write sql code directly in rmarkdown as well
tbl(con, "bio_table") %>% 
  filter(state == "NC") %>% 
  select(name, city, capacity)


```

# Data Cleaning {.tabset}

## Print Data

**How do you take a look at your data in R?**

```{r}
bio %>% 
  glimpse()
```

```{r}
giving %>% 
  glimpse()
```

**Does anything look off about the giving data?**

```{r}
clean_names(giving)

giving <- clean_names(giving)

```

## Missing Values

### How missing values work in R

R uses the `NA` code for missing values. You can test if a value is missing using the `is.na()` function. 

**How many missing values are there in the deceased variable?**

```{r missing_data}

is.na(bio$deceased)[1:100] 

sum(is.na(bio$deceased))

bio %>% 
  summarise(deceased_na = sum(is.na(deceased)))

bio %>% 
  summarise_all(funs(sum(is.na(.))))

giving %>% 
  summarise_all(funs(sum(is.na(.))))
```

**Which records are missing zip, state, lat, and lon?**

```{r}

bio %>% 
  filter(is.na(zip)) %>% 
  glimpse()

```

The `zipcode` package can be used to get lat/lon coordinates for each zipcode's centroid in the US. This data is also available [here](http://federalgovernmentzipcodes.us/free-zipcode-database-Primary.csv).


## Data Cleaning

You can treat character, numeric, and factor variables seperately using variations of the `select` function. 

```{r data_cleaning}
bio %>% 
  select_if(is.numeric)

bio %>% 
  select_if(is.character)
```

**Which variable is should not be a character?**

```{r}
bio <-
  bio %>% 
  mutate(zip = as.numeric(zip))
```


**How might we recode the missing values for the deceased variable?** 
```{r}
bio <-
  bio %>% 
  mutate(deceased_missing = ifelse(is.na(deceased), "Y", "N"),
         deceased = ifelse(is.na(deceased), "N", deceased)) 

```

**Are certain capacity sources missing capacity information?**
```{r}
# capacity source 
bio %>% 
  count(capacity, capacity_source)

bio %>%
  filter(is.na(capacity_source)) %>% 
  count(capacity, capacity_source)

```

**Let's take a closer look at the birthday variable. What do you notice when we sort all birthdays in order?**
```{r}
# birthdays - let's sort all the birthdays in order 
bio %>% 
  select(birthday, deceased) %>% 
  arrange(birthday)

# let's take a look at the distribution of birthdays
bio %>% 
  select(birthday) %>%  
  ggplot(aes(x = birthday)) +
  geom_histogram() 

# let's clean up what appears to be a missing value indicator 
bio <-
  bio %>%
  mutate(birthday = if_else(birthday == as.Date("1/1/1900", "%m/%d/%Y"),
                            as.Date(NA),
                            birthday))

# let's take another look
bio %>% 
  select(birthday) %>%  
  ggplot(aes(x = birthday)) +
  geom_histogram() 
```


# Data Exploration {.tabset}

## Small Multiples
```{r prelim_data_viz}
# bio table - character variables bar plots
bio %>% 
  select_if(is.character) %>% 
  select(-name, -city) %>% 
  gather("variable", "value") %>% 
  ggplot(aes(x = value)) +
  geom_bar() +
  facet_wrap(~variable, scales = "free", nrow = 2) +
  theme(axis.text.y = element_text(size = 6)) +
  coord_flip()
```

**What looks strange?**

## More Cleaning
```{r}
# clean capacity ratings
sort(unique(bio$capacity))

# demo multiple cursors

#  [1] ">$1k"
#  [2] "$100k - $250k"
#  [3] "$10k - $25k"
#  [4] "$10M - $25M"
#  [5] "$1k - $2.5k"
#  [6] "$1M - $2.5M"
#  [7] "$2.5k - $5k"
#  [8] "$250k - $500k"
#  [9] "$25k - $50k"
# [10] "$25M - $50M"
# [11] "$500k - $750k"
# [12] "$50k - $75K"
# [13] "$5k - $10k"
# [14] "$5M - $10M"
# [15] "$750k - $1M"
# [16] "$75k - $100k" 

bio <-
  bio %>% 
  mutate(capacity = factor(capacity, levels = c(">$1k",
                                                "$1k - $2.5k",
                                                "$2.5k - $5k",
                                                "$5k - $10k",
                                                "$10k - $25k",
                                                "$25k - $50k",
                                                "$50k - $75K",
                                                "$75k - $100k",
                                                "$100k - $250k",
                                                "$250k - $500k",
                                                "$500k - $750k",
                                                "$750k - $1M",
                                                "$1M - $2.5M",
                                                "2.5M - $5M",
                                                "$5M - $10M",
                                                "$10M - $25M",
                                                "$25M - $50M")))

# let's take another look at those capacities
bio %>% 
  select(capacity) %>% 
  ggplot(aes(x = capacity)) +
  geom_bar() +
  coord_flip()
```

## Closer Look at State 

```{r}
# state
bio %>% 
  filter(!is.na(state)) %>% 
  count(state) %>% 
  arrange(desc(n)) %>% 
  slice(1:10) %>% 
  ggplot(aes(x = reorder(state, n),  y = n)) +
  geom_bar(stat = "identity", fill = "#027854") +
  coord_flip() +
  ggthemes::theme_tufte() +
  labs(y = "Number of Prospects", 
       x = "Primary Residence State",
       title = "Prospects by State")

```

*Is this right? Do we need to exclude some prospects?*

```{r}
# state
state_plot <- 
bio %>% 
  filter(!is.na(state),
         deceased == "N",
         !duplicated(household_id)) %>% 
  count(state) %>% 
  arrange(desc(n)) %>% 
  slice(1:10) %>% 
  ggplot(aes(x = reorder(state, n),  y = n)) +
  geom_bar(stat = "identity", fill = "#027854") +
  coord_flip() +
  ggthemes::theme_tufte() +
  labs(y = "Number of Prospects", 
       x = "Primary Residence State",
       title = "Prospects by State")

ggplotly(state_plot)

```

## Giving Data

**How would we plot the distribution of gift dates (i.e., the number of gifts per day)?**
```{r}
# gifts per day
giving %>% 
  filter(credit_type == "Hard-Credit") %>% 
  ggplot(aes(x = gift_date)) +
  geom_histogram()
```

**How about the distribution of gift amounts?**
```{r}
giving %>% 
  filter(credit_type == "Hard-Credit") %>% 
  ggplot(aes(x = gift_amt)) +
  geom_histogram()

giving %>% 
  filter(credit_type == "Hard-Credit",
         gift_amt < 1000000) %>% 
  ggplot(aes(x = gift_amt)) +
  geom_histogram()

giving %>% 
  filter(credit_type == "Hard-Credit",
         gift_amt < 100000) %>% 
  ggplot(aes(x = gift_amt)) +
  geom_histogram()

```

# How does our fundraising progress compare to previous fiscal years? 

*What's our first step?*

```{r}
giving <-
  giving %>% 
  mutate(fy = ifelse(month(gift_date) >= 7, 
                     year(gift_date) +1, 
                     year(gift_date)))

giving %>% 
  count(fy)

giving %>% 
  filter(credit_type == "Hard-Credit") %>% 
  group_by(fy) %>% 
  summarise(total_giving = dollar(sum(gift_amt)))


  
```

*Is this it? What else might we need to account for?* 


```{r}
calculateFY <- function(date = Sys.Date(), date.format = "%Y-%m-%d", ytd = FALSE, fiscal.year = 2020){ 
  
  date <- as.Date(date, date.format)
  
  fy.date <- 
    ifelse(month(date) %in% c(1:6), 
           year(date),
           year(date) + 1)
  
  if(ytd == TRUE){
    
    fy <- fiscal.year
    
    end.this.fy  <- as.Date(paste0("6/30/", fy), format = "%m/%d/%Y")
    
    days.left.this.fy <- end.this.fy - Sys.Date()
    
    end.date.fy  <- as.Date(paste0("6/30/", fy.date), format = "%m/%d/%Y")
    
    days.left.date.fy <- end.date.fy - date
    
    if(days.left.date.fy >= days.left.this.fy){
      
      return(fy.date)
    
    }else{
        
      return(NA)
      
      }
    
  }else{
    
    return(fy.date)
    
  }
  
}
# giving$fy <- unlist(lapply(giving$gift_date, function(x) suppressWarnings(calculateFY(x, ytd = TRUE))))

ytd_table <- tibble(
  gift_date = seq(min(giving$gift_date), max(giving$gift_date), by = "day"),
)

ytd_table$fy_ytd <- unlist(lapply(ytd_table$gift_date, function(x) suppressWarnings(calculateFY(x, ytd = TRUE))))

giving %>% 
  left_join(ytd_table) %>% 
  filter(!is.na(fy_ytd)) %>% 
  filter(credit_type == "Hard-Credit") %>% 
  group_by(fy_ytd) %>% 
  summarise(total_giving = dollar(sum(gift_amt)))


```

There is a fundraising R package in development that may help and is available [here](https://github.com/crazybilly/fundRaising). 

# Which prospects should we rate next? 

*What might our first step be?*

```{r}
# calculate annual and total giving
# see who is not rated or rated low
giving_by_household_and_fy <- 
  giving %>% 
  group_by(household_id, fy) %>% 
  summarise(giving = sum(gift_amt)) %>% 
  spread(fy, giving, sep = "") %>% 
  ungroup() %>% 
  mutate(total_giving = rowSums(select(., contains("fy")), na.rm = TRUE))

sum(duplicated(giving_by_household_and_fy$household_id))

bio_with_household_giving <- 
  bio %>% 
  filter(!duplicated(household_id)) %>% 
  left_join(giving_by_household_and_fy)


bio_with_household_giving %>% 
  filter(capacity_source %in% c(NA, "screening")) %>% 
  filter(total_giving > 10000) %>% 
  filter(!is.na(fy2019)) %>% 
  arrange(desc(total_giving)) %>% 
  select(name, capacity, capacity_source, contains("fy"), total_giving) %>% 
  datatable(rownames = FALSE) %>% 
  formatCurrency(columns = c(3:10), digits = 0)

```

# Where should we host a fundraising event? 

*What might our first step be?*

```{r}
bio_with_household_giving %>% 
  filter(total_giving > 10000) %>% 
  filter(!is.na(fy2019)) %>% 
  leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(clusterOptions = markerClusterOptions(),
                   label = ~paste0(name, ": ", scales::dollar(total_giving)))

```








